\chapter{State Transition and Uncertainty Propagation}
\label{ch:state_transition}

\section{Introduction}

The \textbf{state transition matrix} (STM) is fundamental to orbit determination, tracking small perturbations in orbital motion and enabling both linear and non-linear uncertainty propagation. This chapter develops the mathematical theory of the STM and describes the implemented tools in AstDyn for propagating covariance matrices and performing Monte Carlo analysis.

\section{Mathematical Foundation}

\subsection{Linearization of Dynamics}

Consider the general orbital dynamics:

\begin{equation}
    \dot{\mathbf{y}} = \mathbf{f}(t, \mathbf{y})
\end{equation}

where $\mathbf{y} = [\mathbf{r}, \mathbf{v}]^T$ is the 6-dimensional state vector.

For a reference trajectory $\mathbf{y}_{\text{ref}}(t)$ and a perturbed trajectory $\mathbf{y}(t)$, define:

\begin{equation}
    \delta\mathbf{y}(t) = \mathbf{y}(t) - \mathbf{y}_{\text{ref}}(t)
\end{equation}

\subsection{Variational Equations}

Assuming small perturbations, we linearize:

\begin{equation}
    \delta\dot{\mathbf{y}} = \frac{\partial\mathbf{f}}{\partial\mathbf{y}}\bigg|_{\mathbf{y}_{\text{ref}}} \delta\mathbf{y} = \mathbf{A}(t) \delta\mathbf{y}
\end{equation}

where $\mathbf{A}(t)$ is the $6 \times 6$ Jacobian matrix:

\begin{equation}
    \mathbf{A} = \begin{bmatrix}
        \frac{\partial\mathbf{f}_r}{\partial\mathbf{r}} & \frac{\partial\mathbf{f}_r}{\partial\mathbf{v}} \\[8pt]
        \frac{\partial\mathbf{f}_v}{\partial\mathbf{r}} & \frac{\partial\mathbf{f}_v}{\partial\mathbf{v}}
    \end{bmatrix} = \begin{bmatrix}
        \mathbf{0}_{3\times3} & \mathbf{I}_{3\times3} \\[8pt]
        \frac{\partial\mathbf{a}}{\partial\mathbf{r}} & \frac{\partial\mathbf{a}}{\partial\mathbf{v}}
    \end{bmatrix}
\end{equation}

\subsection{State Transition Matrix Definition}

The \textbf{state transition matrix} $\Phi(t, t_0)$ is the solution to:

\begin{equation}
    \frac{d\Phi}{dt} = \mathbf{A}(t)\Phi(t, t_0), \quad \Phi(t_0, t_0) = \mathbf{I}_{6\times6}
\end{equation}

It relates state perturbations at different times:

\begin{equation}
    \delta\mathbf{y}(t) = \Phi(t, t_0) \delta\mathbf{y}(t_0)
\end{equation}

\subsection{Properties}

The STM has important properties:

\begin{enumerate}
    \item \textbf{Identity at $t_0$}: $\Phi(t_0, t_0) = \mathbf{I}$
    \item \textbf{Composition}: $\Phi(t_2, t_0) = \Phi(t_2, t_1)\Phi(t_1, t_0)$
    \item \textbf{Inverse}: $\Phi(t_0, t) = \Phi^{-1}(t, t_0)$
    \item \textbf{Determinant}: $\det[\Phi(t, t_0)] = \exp\left[\int_{t_0}^t \text{tr}(\mathbf{A}(\tau))d\tau\right]$
\end{enumerate}

For conservative systems (Hamiltonian), the STM is symplectic: $\Phi^T\mathbf{J}\Phi = \mathbf{J}$ where $\mathbf{J}$ is the symplectic matrix.

\section{Jacobian Matrix Computation}

\subsection{Two-Body Problem}

For the unperturbed Kepler problem:

\begin{equation}
    \mathbf{a} = -\frac{\mu}{r^3}\mathbf{r}
\end{equation}

The acceleration partials are:

\begin{equation}
    \frac{\partial\mathbf{a}}{\partial\mathbf{r}} = -\frac{\mu}{r^3}\left[\mathbf{I} - 3\frac{\mathbf{r}\mathbf{r}^T}{r^2}\right]
\end{equation}

\begin{equation}
    \frac{\partial\mathbf{a}}{\partial\mathbf{v}} = \mathbf{0}_{3\times3}
\end{equation}

Thus:

\begin{equation}
    \mathbf{A}_{\text{2-body}} = \begin{bmatrix}
        \mathbf{0} & \mathbf{I} \\[8pt]
        -\frac{\mu}{r^3}\left[\mathbf{I} - 3\frac{\mathbf{r}\mathbf{r}^T}{r^2}\right] & \mathbf{0}
    \end{bmatrix}
\end{equation}

\subsection{N-Body Perturbations}

For planetary perturbations, the acceleration is:

\begin{equation}
    \mathbf{a}_p = \mu_p \left[\frac{\mathbf{r}_p - \mathbf{r}}{|\mathbf{r}_p - \mathbf{r}|^3} - \frac{\mathbf{r}_p}{r_p^3}\right]
\end{equation}

The partial derivative with respect to position:

\begin{equation}
    \frac{\partial\mathbf{a}_p}{\partial\mathbf{r}} = -\frac{\mu_p}{d^3}\left[\mathbf{I} - 3\frac{\mathbf{d}\mathbf{d}^T}{d^2}\right]
\end{equation}

where $\mathbf{d} = \mathbf{r}_p - \mathbf{r}$ and $d = |\mathbf{d}|$.

\subsection{Relativistic Corrections}

The post-Newtonian acceleration includes velocity-dependent terms:

\begin{equation}
    \mathbf{a}_{\text{GR}} = \frac{\mu}{c^2 r^3}\left[4\frac{\mu}{r}\mathbf{r} - v^2\mathbf{r} + 4(\mathbf{r} \cdot \mathbf{v})\mathbf{v}\right]
\end{equation}

Both $\partial\mathbf{a}_{\text{GR}}/\partial\mathbf{r}$ and $\partial\mathbf{a}_{\text{GR}}/\partial\mathbf{v}$ are non-zero.

For position:

\begin{equation}
    \frac{\partial\mathbf{a}_{\text{GR}}}{\partial\mathbf{r}} = \frac{\mu}{c^2 r^3}\left[-v^2\mathbf{I} + 4(\mathbf{v}\mathbf{v}^T) + \text{(higher order terms)}\right]
\end{equation}

For velocity:

\begin{equation}
    \frac{\partial\mathbf{a}_{\text{GR}}}{\partial\mathbf{v}} = \frac{\mu}{c^2 r^3}\left[-2v\mathbf{r}\mathbf{v}^T + 4\mathbf{v}\mathbf{r}^T + 4(\mathbf{r} \cdot \mathbf{v})\mathbf{I}\right]
\end{equation}

\subsection{Solar Radiation Pressure}

For SRP with constant area-to-mass ratio:

\begin{equation}
    \mathbf{a}_{\text{SRP}} = P_\odot \frac{A}{m} C_R \left(\frac{r_0}{r}\right)^2 \hat{\mathbf{r}}
\end{equation}

The partial is:

\begin{equation}
    \frac{\partial\mathbf{a}_{\text{SRP}}}{\partial\mathbf{r}} = P_\odot \frac{A}{m} C_R r_0^2 \left[\frac{\mathbf{I}}{r^3} - 3\frac{\mathbf{r}\mathbf{r}^T}{r^5}\right]
\end{equation}

\section{Numerical Computation}

\subsection{Augmented State Vector}

To compute the STM numerically, augment the state vector:

\begin{equation}
    \tilde{\mathbf{y}} = \begin{bmatrix} \mathbf{y} \\ \text{vec}(\Phi) \end{bmatrix} \in \mathbb{R}^{42}
\end{equation}

where $\text{vec}(\Phi)$ stacks the 36 elements of $\Phi$ column-wise.

\subsection{Augmented Dynamics}

The augmented system is:

\begin{equation}
    \frac{d\tilde{\mathbf{y}}}{dt} = \begin{bmatrix} \mathbf{f}(\mathbf{y}) \\ \text{vec}(\mathbf{A}(\mathbf{y})\Phi) \end{bmatrix}
\end{equation}

In practice, we integrate:
\begin{itemize}
    \item 6 equations for the state $\mathbf{y}$
    \item 36 equations for the STM elements
    \item Total: 42 coupled ODEs
\end{itemize}

\subsection{Implementation in AstDyn: GRKNIntegrator}

The primary way to propagate the STM in AstDyn is using the \texttt{GRKNIntegrator}, a specialized Runge-Kutta-Nyström integrator that can simultaneously evolve the state and its variational equations.

\begin{lstlisting}[language=C++,caption={STM propagation using GRKNIntegrator}]
#include "astdyn/propagation/GRKNIntegrator.hpp"

using namespace astdyn::propagation;

// 1. Setup the integrator with physical parameters
GRKNIntegrator grkn(
    1e-12,    // tolerance
    0.1,      // initial step [days]
    1e-8,     // min step
    100.0,    // max step
    mu_sun,   // GM
    j2_sun    // J2 (optional)
);

// 2. Prepare the augmented state (42 elements: 6 state + 36 STM)
Eigen::VectorXd y0 = Eigen::VectorXd::Zero(42);
y0.head<6>() = initial_cartesian_au_aud; // [x,y,z,vx,vy,vz] in AU, AU/d
// Initial STM is local identity
Eigen::Map<Eigen::Matrix<double, 6, 6>>(y0.data() + 6) = 
    Eigen::Matrix<double, 6, 6>::Identity();

// 3. Propagate
auto y_final = grkn.integrate(derivative_func, y0, t0, tf);

// 4. Extract results
Eigen::Vector3d rf = y_final.head<3>();
Eigen::Vector3d vf = y_final.segment<3>(3);
Eigen::Matrix<double, 6, 6> Phi = 
    Eigen::Map<Eigen::Matrix<double, 6, 6>>(y_final.data() + 6);

std::cout << "STM determinant: " << Phi.determinant() << "\n";
\end{lstlisting}

\subsection{Computational Cost}

STM computation increases computational cost:

\begin{table}[htbp]
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Computation} & \textbf{State Equations} & \textbf{CPU Time Factor} \\
\midrule
State only & 6 & 1.0$\times$ \\
State + STM & 42 & 5-7$\times$ \\
State + STM + sensitivity & 42 + 6$N_p$ & 10-15$\times$ \\
\bottomrule
\end{tabular}
\caption{Computational cost of STM propagation. $N_p$ is number of parameters.}
\label{tab:stm_cost}
\end{table}

\section{Applications}

\subsection{Orbit Determination}

In differential correction (least squares orbit fitting), we need:

\begin{equation}
    \frac{\partial\mathbf{y}(t_{\text{obs}})}{\partial\mathbf{y}(t_0)} = \Phi(t_{\text{obs}}, t_0)
\end{equation}

This relates observations to initial conditions, enabling iterative orbit refinement.

\subsection{Covariance Propagation}

Given initial covariance $\mathbf{P}_0$, the covariance at time $t$ is:

\begin{equation}
    \mathbf{P}(t) = \Phi(t, t_0) \mathbf{P}_0 \Phi^T(t, t_0)
\end{equation}

This quantifies uncertainty growth over time.

Example:
\begin{lstlisting}[language=C++,caption={Covariance propagation}]
Matrix6d P0 = initial_covariance();  // km^2, (km/s)^2
Matrix6d Phi = result.stm;

Matrix6d Pf = Phi * P0 * Phi.transpose();

// Position uncertainty at final time
Vector3d sigma_pos = Pf.block<3,3>(0,0).diagonal().cwiseSqrt();
std::cout << "Position uncertainty: " 
          << sigma_pos.transpose() << " km\n";
\end{lstlisting}

\subsection{Sensitivity Analysis}

The STM reveals how perturbations in initial conditions affect future states:

\begin{equation}
    \frac{\partial r(t)}{\partial r_0} = \Phi_{11}(t, t_0), \quad
    \frac{\partial r(t)}{\partial v_0} = \Phi_{12}(t, t_0)
\end{equation}

These are the upper-left and upper-right $3 \times 3$ blocks of $\Phi$.

\subsection{Maneuver Optimization}

For spacecraft trajectory design, the STM helps compute:
\begin{itemize}
    \item Targeting matrices (where to aim to hit a target)
    \item $\Delta v$ requirements
    \item Sensitivity to execution errors
\end{itemize}

\section{Analytical vs Numerical STM}

\subsection{Analytical STM for Keplerian Motion}

For the unperturbed two-body problem, closed-form solutions exist. The STM can be expressed in terms of orbital elements and their derivatives.

Advantages:
\begin{itemize}
    \item Exact (no numerical error)
    \item Fast to evaluate
    \item Valid for long time spans
\end{itemize}

Disadvantages:
\begin{itemize}
    \item Complex formulas (especially near singularities)
    \item Doesn't include perturbations
    \item Limited practical use
\end{itemize}

\subsection{Numerical STM}

Integrating the variational equations numerically:

Advantages:
\begin{itemize}
    \item Handles arbitrary force models
    \item Straightforward implementation
    \item Includes all perturbations
\end{itemize}

Disadvantages:
\begin{itemize}
    \item Numerical error accumulation
    \item 7$\times$ slower than state-only propagation
    \item Ill-conditioning for long arcs
\end{itemize}

\subsection{Hybrid Approaches}

For some applications, use:
\begin{enumerate}
    \item Analytical STM for Keplerian part
    \item Numerical perturbation corrections
    \item State transition composition
\end{enumerate}

\section{Numerical Stability}

\subsection{Conditioning Issues}

The STM becomes ill-conditioned for:
\begin{itemize}
    \item Long propagation times ($>$ several orbital periods)
    \item High eccentricity orbits
    \item Nearly rectilinear motion
\end{itemize}

Condition number growth:

\begin{equation}
    \kappa(\Phi) \approx \exp\left(\lambda_{\max} \Delta t\right)
\end{equation}

where $\lambda_{\max}$ is the largest Lyapunov exponent.

\subsection{Mitigation Strategies}

\textbf{1. Relinearization}

Instead of propagating from $t_0$ to $t_f$, divide into segments:

\begin{equation}
    \Phi(t_f, t_0) = \Phi(t_f, t_2) \Phi(t_2, t_1) \Phi(t_1, t_0)
\end{equation}

Each segment has better conditioning.

\textbf{2. State transition in orbital elements}

Instead of Cartesian STM, use:

\begin{equation}
    \frac{\partial\mathbf{e}(t)}{\partial\mathbf{e}(t_0)}
\end{equation}

where $\mathbf{e} = [a, e, i, \Omega, \omega, M]$ are orbital elements.

\textbf{3. Regularization}

Use regularized coordinates (Kustaanheimo-Stiefel, Sperling-Burdet) that are better behaved near periapsis.

\section{Practical Example}

\subsection{Target Tracking}

Track uncertainty in asteroid position for impact assessment:

\begin{lstlisting}[language=C++,caption={Asteroid uncertainty propagation}]
// Initial state from orbit determination
Vector6d y0 = {1.1, 0.2, 0.05, -0.01, 0.03, 0.0};  // AU, AU/day

// Initial covariance (from least squares fit)
Matrix6d P0 = Matrix6d::Zero();
P0.diagonal() << 1e-8, 1e-8, 1e-9,  // pos: 1500 km
                 1e-11, 1e-11, 1e-12;  // vel: 0.15 m/s

ForceModel forces;
forces.enable_planets({"Earth", "Jupiter", "Venus", "Mars"});

Propagator prop(forces);
prop.enable_stm(true);

// Propagate 10 years
double t0 = 60000.0;
double tf = t0 + 3652.5;  // 10 years

auto result = prop.propagate_with_stm(y0, t0, tf);

// Compute uncertainty at future time
Matrix6d Pf = result.stm * P0 * result.stm.transpose();

// Position uncertainty (3-sigma)
Vector3d sigma_3 = 3.0 * Pf.block<3,3>(0,0).diagonal().cwiseSqrt();
std::cout << "Position uncertainty (3-sigma): \n";
std::cout << sigma_3.transpose() * AU_TO_KM << " km\n";

// Check for Earth close approach
Vector6d earth_state = ephemeris.get_planet("Earth", tf);
Vector3d rel_pos = result.state.head<3>() - earth_state.head<3>();
double distance = rel_pos.norm() * AU_TO_KM;

std::cout << "Distance to Earth: " << distance << " km\n";
std::cout << "Impact probability (Gaussian): ";
if (distance < 3.0 * sigma_3.norm() * AU_TO_KM) {
    std::cout << "NON-ZERO - further analysis required\n";
} else {
    std::cout << "Negligible\n";
}
\end{lstlisting}

\subsection{Observation Planning}

Determine optimal observation times to reduce uncertainty:

\begin{lstlisting}[language=C++,caption={Observation planning}]
// Propagate with STM to multiple observation epochs
std::vector<double> obs_times = {t0 + 30, t0 + 60, t0 + 90};

for (double t_obs : obs_times) {
    auto result = prop.propagate_with_stm(y0, t0, t_obs);
    Matrix6d P = result.stm * P0 * result.stm.transpose();
    
    // RA/Dec uncertainty from position uncertainty
    Vector3d r = result.state.head<3>();
    double dec = std::asin(r(2) / r.norm());
    double ra = std::atan2(r(1), r(0));
    
    // Simple approximation (full calculation uses observation partials)
    double sigma_ra = P(0,0) / (r.norm() * std::cos(dec));
    double sigma_dec = P(2,2) / r.norm();
    
    std::cout << "Epoch " << t_obs << ": "
              << "sigma_RA = " << sigma_ra * RAD_TO_ARCSEC << " arcsec, "
              << "sigma_Dec = " << sigma_dec * RAD_TO_ARCSEC << " arcsec\n";
}
\end{lstlisting}

\section{Parameter Sensitivity}

\subsection{Extended State Vector}

To study sensitivity to dynamical parameters (e.g., $\mu$, $C_R$, asteroid masses), augment the state:

\begin{equation}
    \tilde{\mathbf{y}} = \begin{bmatrix} \mathbf{y} \\ \mathbf{p} \end{bmatrix}
\end{equation}

where $\mathbf{p}$ are parameters. Then:

\begin{equation}
    \frac{d}{dt}\begin{bmatrix} \mathbf{y} \\ \mathbf{p} \end{bmatrix} = \begin{bmatrix} \mathbf{f}(\mathbf{y}, \mathbf{p}) \\ \mathbf{0} \end{bmatrix}
\end{equation}

The extended STM includes $\partial\mathbf{y}/\partial\mathbf{p}$.

\subsection{Sensitivity Matrices}

Define sensitivity matrix:

\begin{equation}
    \mathbf{S}(t) = \frac{\partial\mathbf{y}(t)}{\partial\mathbf{p}}
\end{equation}

It satisfies:

\begin{equation}
    \frac{d\mathbf{S}}{dt} = \mathbf{A}(t)\mathbf{S} + \frac{\partial\mathbf{f}}{\partial\mathbf{p}}
\end{equation}

This reveals how orbital motion depends on physical parameters.

\section{Uncertainty Propagation}

\subsection{Linear Covariance Propagation}

For systems where the deviations from the nominal trajectory remain small, the uncertainty can be represented by a $6 \times 6$ covariance matrix $\mathbf{P}$. AstDyn implements linear propagation via the \texttt{CovariancePropagator} class, which uses the state transition matrix $\mathbf{\Phi}$ and an optional process noise matrix $\mathbf{Q}$.

\subsubsection{Mathematical Formulation}

The covariance matrix at a future time $t_f$ is computed as:
\begin{equation}
    \mathbf{P}(t_f) = \mathbf{\Phi}(t_f, t_0) \mathbf{P}(t_0) \mathbf{\Phi}^T(t_f, t_0) + \mathbf{Q}
\end{equation}
where $\mathbf{Q}$ accounts for unmodeled dynamical noise or integration errors accumulation.

\subsubsection{Numerical STM via Finite Differences}

While high-precision propagation can use the variational equations (Chapter 10.4), the \texttt{CovariancePropagator} also supports a robust numerical STM established via symmetric finite differences:
\begin{equation}
    \mathbf{\Phi}_{ij} \approx \frac{\mathbf{y}_i(t_f, \mathbf{y}_0 + \Delta\mathbf{y}_j) - \mathbf{y}_i(t_f, \mathbf{y}_0 - \Delta\mathbf{y}_j)}{2 \Delta y_j}
\end{equation}
This approach is particularly stable for complex force models where analytical partials might be unavailable or computationally prohibitive.

\begin{lstlisting}[language=C++,caption={Linear Covariance Propagation Example}]
#include "astdyn/propagation/CovariancePropagator.hpp"

// 1. Setup the propagator
auto cov_prop = CovariancePropagator::make(propagator);

// 2. Propagate mean state and covariance
auto result = cov_prop.propagate_with_covariance(
    initial_state,
    initial_covariance,
    target_epoch,
    process_noise // Q matrix
);

std::cout << "Final uncertainty (1-sigma): " 
          << result.covariance.sigma_pos_km() << " km\n";
\end{lstlisting}

\subsection{Non-linear Monte Carlo Uncertainty}

When the dynamics are highly non-linear (e.g., during close planetary encounters) or the initial uncertainty is large, linear propagation via the STM may become inaccurate. AstDyn provides a high-performance \texttt{MonteCarloUncertainty} module for ensemble propagation.

\subsubsection{Algorithm}

The Monte Carlo engine follows a three-step process:
\begin{enumerate}
    \item \textbf{Sampling}: Generate an ensemble of $N$ states by sampling from the multivariate normal distribution $\mathcal{N}(\mathbf{y}_{\text{nominal}}, \mathbf{P}_0)$ using Cholesky decomposition of the covariance matrix.
    \item \textbf{Propagation}: Parallel propagation of all $N$ samples to the target epoch using the system's numerical integrators.
    \item \textbf{Statistics}: Re-computation of the mean state and covariance matrix from the final ensemble.
\end{enumerate}

\subsubsection{Parallel Execution}

To handle the significant computational cost of propagating thousands of samples, the \texttt{MonteCarloUncertainty} module leverages OpenMP multi-threading, achieving near-linear speedup on multi-core workstations.

\begin{lstlisting}[language=C++,caption={Monte Carlo Uncertainty Analysis}]
#include "astdyn/propagation/MonteCarloUncertainty.hpp"

auto mc = MonteCarloUncertainty::make(propagator);

// 1. Generate 10,000 samples
auto ensemble = mc.generate_samples(nominal_state, P0, 10000);

// 2. Propagate in parallel (OpenMP)
auto final_states = mc.propagate_ensemble(ensemble, target_epoch, *propagator);

// 3. Compute statistics
auto stats = mc.compute_statistics(final_states);
std::cout << "MC Mean position: " << stats.mean.position() << "\n";
std::cout << "MC Position RMS: " << stats.position_rms_km << " km\n";
\end{lstlisting}

\section{Summary}

Key concepts about state transition and uncertainty propagation:

\begin{enumerate}
    \item The \textbf{STM} $\Phi(t, t_0)$ propagates small perturbations linearly.
    \item \textbf{Linear Covariance Propagation} uses $\mathbf{\Phi} \mathbf{P} \mathbf{\Phi}^T + \mathbf{Q}$ for efficient uncertainty forecasting.
    \item \textbf{Monte Carlo methods} handle non-linear dynamics and large initial errors via ensemble sampling.
    \item AstDyn provides specialized classes (\texttt{CovariancePropagator}, \texttt{MonteCarloUncertainty}) optimized for both accuracy and performance.
\end{enumerate}

Understanding these tools is essential for:
\begin{itemize}
    \item Impact probability assessment and collision avoidance.
    \item Observation planning and survey optimization.
    \item Validating the limits of linear approximation in complex dynamical environments.
\end{itemize}
